---
title: "Exercise 4: Applied Statistics"
format: html
editor: source
author: "HeaDS Data Science Lab, University of Copenhagen"
project:
  type: website
  output-dir: ../docs
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This exercise focuses on **Gaussian linear models**, where data points are assumed to be independent, normally distributed, and have equal variance. Such models are fitted in R using `lm()`, e.g.

Later sections also introduce **logistic regression** (for binary outcomes) and **linear mixed models** (for grouped or hierarchical data).

The purpose of this exercise is to show how to fit and work with statistical models in R. This means that the analyses are not necessarily those that should be used in a real-life analysis of the data.

Start with the **Data** part.

### Data and Packages

```{r}
library(tidyverse)
library(ggfortify)
```

The dataset to be used concerns risk factors for low infant birth weight and is available in the **MASS** package. It includes 189 births with the following variables:

1.  Use the commands below to load the MASS package and get the help page for the data. The help page will appear in the lower right window of RStudio. Read about the data; in particular about the variables `age`, `lwt`, `ftv`, `smoke`, and `bwt`.

```{r}
library(MASS)
data(birthwt) # Calls the data frame called birthwt (Promise)
head(birthwt) # Loads the data in Environment and shows the head
```

2.   `birthwt` is not a tibble (the modern tidyverse data format). Convert it to a tibble.

```{r}

```

3.  The variables `smoke` (mother’s smoking habits) and `ftv` (number of first-trimester physician visits) are coded as numeric variables.\
    Convert them to **factors** so they are treated as categorical variables in your analyses:

```{r}

```

4.  Make boxplot of birth weight `bwt` according the number of visits to physician `ftvFac` during the first trimester. Split by smoking status.

```{r}

```

5.  Make a scatterplot with `lwt` on the $x$-axis and `bwt` on the $y$-axis. Modify the plot such that points are colored according to `smoke` and size shows `ftv.`

```{r}

```

6.  The `ftv` is a numerical variable, but we would also like a categorical version of it called `visits` , with groups corresponding to zero visits, one visit, and two or more visits, respectively.

::: {.callout-tip collapse="true"}
## Hint

Use `mutate` and take a look at `fct_collapse` function.
:::

```{r}

```

7.  How many women had more than one physician visit during the first trimester?

### Regression

The term *regression* is usually used for models where all explanatory variables are numerical.

If you are not too familiar with linear regression, you can have a look at what the `summary()` output means [here](https://feliperego.github.io/blog/2015/10/23/Interpreting-Model-Output-In-R) and for understanding regression `plot()`, have a look [here](https://data.library.virginia.edu/diagnostic-plots/).

8.  Fit a linear regression model with `bwt` as response and `lwt` as covariate, and identify the estimates for the intercept and for the slope. Find the 97.5% confidence interval for the slope parameter by using the `confint()` function.

```{r}

```

9.  Performe model validation. Hint: use autoplot. Does the model seem appropriate for the data?

```{r}
#par(mfrow=c(2,2))
#plot(reg1)

```

10. Fit the multiple linear regression model where you include `lwt` as well as `age` and `ftv` (numerical variables) as covariates. Identify the parameter estimates, and consider what their interpretation is.

```{r}

```

11. Fit the multiple linear regression model again, but now only using data from mothers with a weight less than 160 pounds (`lwt` \< 160). *Hint:* Use the `filter` function and change the `data` argument in the `lm` command.

```{r}

```

### ANOVA

The term **analysis of variance (ANOVA)** is used for models where all explanatory variables are categorical (factors). Before proceeding, make sure that `smoke` is coded as a factor.

12. Fit a **one-way ANOVA** (using `lm()`) where the mean birth weight is allowed to differ between smokers and non-smokers. Find the estimated birth weight for infants from smokers as well as non-smokers. Hint: use `emmeans()`.

```{r}
library(emmeans)    

```

13. Check, is there a statistically significant difference in mean birth weight between infants of smokers and non-smokers?

```{r}

```

14. Fit the oneway ANOVA where you use `visits` as the explanatory variable. Perform an overall F-test for differences across all levels. Find the estimated birth weight for each group, and make the pairwise comparisons between the groups.\
    *Hint* `drop1(, test="F")`. What do you conclude about the relationship between the number of visits and infant birth weight?

```{r}

```

15. Now, consider **both `visits` and `smoke`** as explanatory variables.\
    Since both are categorical, the relevant model is a **two-way ANOVA**. Fit the model without interaction. Inspect the estimated means and group differences. How do you interpret the estimated means? What do they tell you about birth weight by smoking status and visit frequency?

```{r}

```

<!-- -->

16. Fit the twoway ANOVA model *with interaction* replacing `+` by a `*`. Use `drop1()` to test whether the interaction term is statistically significant. Compare it to model without interaction? What do the p-values suggest about whether the effect of smoking depends on the number of visits?

```{r}

```

17. Use `anova()` to formally compare the two models. Are the two models statistically different?\
    What does this imply about including the interaction term?

```{r}

```

18. Are the 2 models statistically different?

## Optional section

### Logistic regression

19. The variable `low` is **1** if birth weight is less than 2500 g, and **0** otherwise.\
    Visualize the data:

```{r}

```

Imagine the actual birth weight `bwt` was **not recorded**, but we still want to know whether the birth weight was low.\
Because the outcome is binary (`low` has two values), we use **logistic regression** to model the probability.

For example, we may consider a model with mothers's weight, smoking status, and age as predictors, just like in the previous question.

20. Fit the model, take a look at `glm` function:

```{r}

```

21. Use `drop1()` to check if each predictor contributes significantly:

    A Note: `test = "F"` is for Gaussian linear models (`lm()`), where F-tests are appropriate.\
    For logistic regression and other GLMs, `test = "Chisq"` is the correct choice.

```{r}

```

```         
**Question:** Based on the p-values, is there evidence that **smoking** affects the probability of low birth weight?
```

22. The fitted model may be used to predict the outcome for new values. The new values are provided in a `newData` data frame with the same names as the covariate in the model.

```{r}
# new data
newData <- tibble(
    low = rep(NA, 5),
    age = c(22, 30, 18, 17, 28),
    lwt = c(140, 155, 130, 97, 165),
    race = c("Black", "Other", "Black", "Other", "White"),
    smoke = as.factor(c(1, 0, 1, 1, 0)),
    ptl = c(1, 0, 0, 0, 0),
    ftv = c(1, 3, 0, 0, 2),
    bwt = c(2800, 4500, 1700, 1500, 3400)
    )
```

23. We have data for **5 new mothers**. Use your fitted logistic model `logreg1` to **predict the probability** of low birthweight and fill the `low` column. Use the `predict()` function to compute the predicted probabilities for the new cases. Make sure to set `type="response"` to get probabilities rather than log-odds.

```{r}

```

24. Add the **predicted probabilities** (red points) and a **connecting line** to your `ggplot`. Hint: add geom_point and geom_line with the newData.

```{r}

```

25. **In how many of the new cases is birthweight predicted to be low?**?
