---
title: "Presentation 3: ggplot2"
format: html
editor: visual
project:
  type: website
  output-dir: ../docs
---

Want to code along? If you haven't already, go to the Data tab of the website and press the **DOWNLOAD PRESENTATIONS** button. This is presentation3.

## Importing libraries and data

```{r, warning=FALSE,message=FALSE}
library(readxl)
library(writexl)
library(tidyverse)
```

## Load data

In the examples below, we use the **downloads** dataset. The dataset is available as an Excel file, which we import using the **readxl** package. Because **ggplot2** is part of the **tidyverse**, we also load the tidyverse packages so we can use its tools for data wrangling and visualization.

We read the *downloads* dataset from the Excel file, and assign it to a `tibble` named *downloads*.

```{r}
downloads <- read_excel("../Data/downloads.xlsx") %>% 
  filter(size > 0)
downloads
```

Here we have filtered out rows where `size` is zero, as they do not represent actual downloads and would distort the visualization.

## ggplot2: The basic concepts

A ggplot object is a structured description of a plot. The function `ggplot()` creates the equivalent of a blank sheet of paper, ready for layers to be added.

The two most important components of any ggplot are:

-   The dataset — typically a `data.frame` or `tibble` containing the variables you want to plot

-   The aesthetics (`aes()`) — a set of mappings that tell ggplot *how* to use the variables: x-axis, y-axis and whether to use color, fill, group, size, shape, etc. to represent groups or magnitudes

```{r}
ggplot(
  downloads,          # dataframe 
  aes(x=machineName)  # x-value
  )    
```

This creates an empty coordinate system with axes defined — essentially a blank canvas.

At this point, `ggplot` has only drawn the axes and labels because we have not yet specified how the data should be visualized.

The next step is to add a geom layer (for example, `geom_bar()`) to actually display the data.

## **Plotting** One Variable

A bar chart is created with `geom_bar()` and is useful for visualizing the distribution of a single categorical variable — for example, `machineName` in this dataset.

`geom_bar()` automatically counts the number of observations in each category and uses those counts as the bar heights. By default, categories on the x-axis are displayed in alphabetical order.

```{r}
ggplot(
  downloads,              # dataframe 
  aes(x=machineName)) +   # categorical variable on x-axis
  geom_bar()              # geometrical object - bar chart
```

Or we can plot a histogram. It shows the distribution of a continuous variable by dividing the data into bins and counting observations in each bin.

```{r}
ggplot(
  downloads,               # dataframe 
  aes(x = log2(size))) +   # continuous variable (log2-transformed)
  geom_histogram()         # geometrical object - histogram
```

## Aesthetics in ggplot2

In ggplot2, aesthetics (`aes()`) map variables in your data to visual properties of the plot. In addition to `x` and `y`, you can map optional aesthetics such as `fill`,  `colour`,  `size`,  `alpha`,  `group`, or `shape`. By changing which aesthetic a variable is mapped to, we can tell ggplot how to display differences between categories.

```{r}
ggplot(
  downloads,
  aes(x = machineName,
      fill = month)) +     # fill by catergorical variable
  geom_bar() 
```

Here, mapping `fill = month` tells ggplot to calculate counts separately for each month and assigns a different fill color to each month and stack the bars.

```{r}
ggplot(
  downloads,
  aes(x = machineName,
      colour = month)) +   # Colour by catergorical variable
  geom_bar() 
```

Here, `colour = month` only changes the outline color of the bars — the interior remains unfilled (or transparent).

## Geometric objects

Geoms (geometric objects) are the core building blocks of ggplot2. They tell ggplot *what kind of plot layer to draw*. Each geom requires data and aesthetics specified inside `aes()`.\
Geoms inherit global aesthetics from the main `ggplot()` call, but you can override them locally inside each geom if needed.

You can adjust how geoms are displayed using various options. For bar charts, position adjustments control how overlapping or grouped data appear. In the previous example, mapping `fill`, `colour`, or `group = month` subdivided each bar by month, producing stacked bars (the default).

You can change this behavior by setting the `position` argument inside `geom_bar()`:

-   `position = "dodge"` → places bars for each group side by side 

-   `position = "fill"` → normalizes bar heights so they represent proportions (percentages).

```{r}
ggplot(
  downloads,
  aes(x = machineName,
      fill = month)) + 
  geom_bar(position = "dodge")  # side by side

```

```{r}
ggplot(
  downloads,
  aes(x = machineName,
      fill = month)) + 
  geom_bar(position = "fill")   # proportions
```

## Plotting Two Variable - A simple Column Chart

To make a bar chart from pre-summarized data, we use `geom_col()` and it requires both x and y values.

Suppose we want to plot the total size of all downloads per machine.\
We first create a summary tibble `dl_sizes` using:

1.  `group_by()` to group the dataset by `machineName`

2.  `summarise()` to calculate the **sum of `size`** within each group

In the example below, we also rescale the download sizes from bytes to megabytes by dividing by 1,000,000.

```{r}
dl_sizes <- downloads %>% 
  group_by(machineName) %>% 
  summarise(size = sum(size))
```

We now plot `machineName` on the x-axis and the total size (in MB) on the y-axis:

```{r}
ggplot(
  dl_sizes, 
  aes(x = machineName, 
      y = size/10^6)) + 
  geom_col()
```

## A Column Chart with Ordered Bars

By default, categorical variables in R are displayed in alphabetical order on the x-axis. Suppose we want the machines in our bar chart to be ordered by increasing total download size instead. This is actually not that easy in R.\
We can do this by converting `machineName` into a factor with levels ordered according to download size.

First, arrange the `dl_sizes` tibble by `size` using `tidyverse` function `arrange`.

```{r}
dl_sizes <- dl_sizes %>% 
  arrange(size)
dl_sizes
```

Next, set the levels of `machineName` according to the arranged order:

```{r}
dl_sizes <- dl_sizes %>% 
  mutate(machineName = factor(machineName, levels = dl_sizes$machineName))

dl_sizes$machineName
```

Now `machineName` is a factor with levels ordered from lowest to highest total download size.

Finally, use the same `ggplot()` code as before to make the plot — now the bars will be ordered by total size rather than alphabetically:

```{r}
ggplot(
  dl_sizes, 
  aes(x = machineName, 
      y = size/10^6)) + 
  geom_col()
```

## A box plot

Boxplots are great to get an overview of continues variables, identifying outliers and comparing distributions across categories.

Here we will for the first time try to save the syntaxical description of a plot in an R variable. This makes it easy to reuse and modify the same plot - for example, to change axes scales or colors - without rewriting the entire code. Let’s try this out and create a boxplot of monthly download sizes for each machine.

Here, we store the bar chart in a variable called `p`. Running this code does not display the plot yet, but a variable named `p` now appears in the Global Environment.

```{r}
p <-                       # save in object p
  ggplot(                  # call ggplot
    downloads,             # data
    aes(x = machineName,   # x and y and fill
        y = size, 
        fill = machineName)) +
  geom_boxplot()           # boxplot

```

Because the data are skewed, we use a logarithmic y-scale to make the plot more interpretable. A logarithmic scale is useful when the data are highly skewed or span several orders of magnitude.

```{r}
p <- p + scale_y_log10()          # Change the scale of the y-axis:
```

Additionally, themes and titles can be added as a layer to any ggplot if you prefer a theme other than the default grey background.

Because the plot is saved in `p`, we can easily modify it by adding more layers:

```{r}
plot1 <- p + 
  theme_minimal() +         # add theme ..................
    # theme_bw()
    # theme_classic()
    # theme_dark()
  labs(x = 'Machines',     # add label ..................
       y = 'Size in Mb',
       title = 'Boxplot')
plot1
```

## Violin plot with `geom_violin`

A violin plot shows the distribution of a continuous variable across different categories, combining the features of a box plot and a density plot.

```{r}
p <-                         # save in object p
  ggplot(                    # call ggplot
    downloads,               # data
    aes(x = machineName,     # x and y and fill
        y = size, 
        fill = machineName)) +
  geom_violin() +            # draw violin plots
  scale_y_log10()            # scale
p

```

We can enhance the plot by overlaying individual data points (with jittering to avoid overplotting), using a minimal theme, and customizing the fill colors:

```{r}
plot2 <- p + 
  geom_jitter(           # overlay geom
    size = 0.1, alpha = 0.05) + 
  theme_minimal() +      # theme
  scale_fill_manual(
    values=c("#2A2D43",  # change colors
             "#91A6FF", 
             "#C7EDE4", 
             "#AFA060", 
             "#AD8350"))
plot2
```

## Scatter plot with `geom_point` and Faceting

A scatter plot is used to explore the relationship between two numerical variables by plotting them on the x- and y-axes. You can visualize a **third** variable (or a condition) by mapping it to the `size` ,`shape` or  `color` aesthetic.

First, we plot **download size vs. time** using `geom_point()` to draw individual data points. We map `time > 500` to the  `size`  aesthetic so downloads taking longer than 500 appear larger. We also map `log10(time)` to  `color`  (so color shows smooth continuous gradient), and use `shape`  to distinguish `machineName`. Because both variables are skewed, we use log scales on both axes with `scale_x_log10()` and `scale_y_log10()`.

```{r}
plot3 <-                       # save in object p
  ggplot(                  # call ggplot
    downloads,             # data
    aes(x = size,          # x and y and color
        y = time, 
        size = time > 500,
        color = log10(time),
        shape = machineName)) +   # add size, color and shape by machineName 
  geom_point() +          # Scaterplot
  scale_colour_continuous(palette = c("darkblue", "gold")) +
  scale_x_log10() +       # scale x and y
  scale_y_log10() + 
  theme_minimal()         # theme

plot3
```

Sometimes adding several aesthetics (like both color and shape) makes a plot too busy, especially if there are many categories. Instead, we can use faceting to split the data into multiple smaller plots - one for each category - making it easier to compare groups.

```{r}
#plot3 + facet_grid(vars(month))
plot3 + facet_wrap(vars(machineName))

```

This wraps the plots into multiple rows and columns automatically, which is helpful when there are many facet levels.

Now you have space to highlight important patterns without overcrowding the plot. For example, faceting makes it easier to see the **relationship (correlation)** between download size and time within each machine, and to spot extreme cases. In particular, we can highlight **very long downloads** (e.g. `time > 500`) by making those points larger, so they stand out clearly.

## Daily summary statistics

Next we want to visualize the number of downloads done each day for each of the 6 machines. We compute the cumulated number of downloads within each of the 6 machines over the days. As before we do this via the following steps:

1.  Using `group_by()` we group the dataset by both *machineName* and *date*.

2.  Using `summarize()` we count the number for within each machine and date.

3.  Using `mutate()` we cumulate the number of downloads over the dates within the machines.

```{r}
daily_downloads <- 
  downloads %>%
    group_by(machineName, date) %>% 
    summarize(dl_count = n()) %>%
    mutate(total_dl_count = cumsum(dl_count))
daily_downloads
```

## Line plot with `geom_line`

The geom called `geom_line()` is used to insert lines, which is ideal for showing trends over time or continuous changes. In the code below you see a first attempt to write R code that visualizes the cumulated total download size over the dates.

We create a plot of the cumulative total download count over time:

```{r}
ggplot(                      # call ggplot
  daily_downloads,           # data
  aes(x = date,              # x and y
      y = total_dl_count)) + # line
  geom_line()

```

If we want a separate line for each machine, we must tell ggplot how to group the data. We can do this by mapping `color = machineName` (and optionally color by machine as well):

```{r}
plot4 <- ggplot(
  daily_downloads, 
  aes(x = date, 
      y = total_dl_count, 
      colour = machineName)) + 
  geom_line() +
  theme_minimal()     # theme
plot4
```

## **Arrange multiple ggplots on the same page**

In R there is a whole infrastructure of extension packages built around ggplot, that facilitate different types of plot modifications, arrangements, and animation.

R package **ggpubr** is one of them and it helps to produce publication-ready plots using ggplot2.

```{r}
# install.packages("ggpubr")
library(ggpubr)

figure <- ggarrange(plot1, plot2, plot3, plot4,
                   align = "hv",
                    labels = c ("A", "B", "C", "D"),
                   ncol = 2, nrow = 2, 
                   legend = "none")
figure
```

```{r}
## Save Tibble and Save Plot:

#write_xlsx(daily_downloads, "daily_downloads.xlsx")

#ggsave(filename = "vionlinPlot.pdf", plot = p2, height = 5, width = 10)

#pdf("vionlinPlot2.pdf", height = 5, width = 10)
#p2
#dev.off()
```

## Conclusion

1.  The **ggplot2** package is a powerful and versatile tool for making plots.

2.  We think, that the generated plots are beautiful.

3.  As far as we know some plots, e.g. using *faceting* (see the exercises), in practice are only producible in **ggplot2**.

4.  Learning the syntax needed for making specific plots is a challenge. The best way (the only way!?) to learn is to practice. You may start by solving the exercise sheet. After you have gotten used to the basic ideas you can find a lot of help on the internet. We remark, that there have been several updates of the **ggplot2** package over the recent years. And some of the old entries that might pop up when you google a ggplot2-issue might be outdated.

5.  Not all things can be made in **ggplot2**! For a geometrical object to be available it need to have a syntaxical description, and it needs to be implemented. Other things require computer-hacks to be made (e.g. using different orderings of categorical variables in faceted plots).
