---
title: "Presentation 1A: Intro to R"
format: html
project:
  type: website
  output-dir: ../docs
---

Want to code along? Go to the Data tab of the website and press the **DOWNLOAD PRESENTATIONS** button. This is presentation1A.

### R as a calculator

Adding

```{r}
1+1
```

Subtracting

```{r}
2-1
```

Multiplication

```{r}
3*3
```

Division

```{r}
15/3
```

Exponentiation

```{r}
10^2
```

Base-2 logarithm

```{r}
log2(20)
```

Base-10 logarithm

```{r}
log10(20)
```

### Define variables

```{r}
a <- 5

b <- 3


a + b 


c <- a + b


c
```

### Check objects in environment

```{r}
ls()
```

### Object Types

Character

```{r}
char1 <- "Hello World!"
char1
class(char1)
```

Numeric

```{r}
num1 <- 5
num1
class(num1)
```

Vector

```{r}
vector1 <- c(1, 2, 3, 4, 5, 'hello', 'world', 7, 1)
vector1
class(vector1)
```

List

```{r}
list1 <- list(1, 2, 3, 4, 5, 'hello', 'world', 7, 1)
list1
class(list1)
```

Vector of numeric

```{r}
vector2 <- c(1, 2, 4, 6, 8, 2, 5, 7)
vector2
```

### Functions

Summing a vector

```{r}
?sum
sum(vector2)
```

Mean of vector

```{r}
?mean
mean(vector2)
```

```{r}
mean(vector2) # mean/average

median(vector2) # median

sd(vector2) # standard deviation

sum(vector2) # sum

min(vector2) # minimum value

max(vector2) # maximum value

length(vector2) # length of vector
```

### Working directory (Exclude/Move)

The term `path` refers to the trajectory you need to follow from the place you are 'located' on your computer to the place you want to work from. When working with Quarto your `working directory (wd)` is always the same locations as your Quarto document (is not true for .R scripts!). The wd becomes important when we start loading data in from other places (presentation 2).

**NB:** These paths are specefic to my computer and the they will be different on your computer.

See current working directory.

```{r, eval=FALSE}
getwd()
```

```         
"/Users/kgx936/Desktop/HeaDS/GitHub_repos/FromExceltoR/Teachers/Presentations"
```

Change working directory.

```{r, eval=FALSE}
setwd("/Users/kgx936/Desktop/HeaDS/GitHub_repos/FromExceltoR")
```

Read excel file from path relative to working directory.

```{r, eval=FALSE}
library(readxl)
read_excel("/Data/climate.xlsx")
```

Read excel file from absolute path.

```{r, eval=FALSE}
read_excel("~/Users/kgx936/Desktop/HeaDS/GitHub_repos/FromExceltoR/Data/climate.xlsx")
```
