---
title: "Data Exercise in R"
author: "HeaDS Data Lab"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
      html_document
      #pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction:

If you don't have your own data to work on you can practice your newly acquired R skills using one of the publicly available datasets we have collected. They are located in the **Exercises/data folder**. Please see below for a brief description of each dataset:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
```

&nbsp;

```{r, echo=FALSE, warning=FALSE, message=FALSE}

datasets <-  tibble(
  Name = c("HealthPrimaryCare", "CrohnsDisease", "StrokeData"),
  Description=c("[Here](https://vincentarelbundock.github.io/Rdatasets/doc/mosaicData/HELPrct.html)", "[Here](https://vincentarelbundock.github.io/Rdatasets/doc/robustbase/CrohnD.html)", "[Here](https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset)"),
  Variables=c(29, 9, 12),
  Observations = c(453, 117, 5110))

knitr::kable(datasets)

```

&nbsp;

The sections which follow serve as inspiration for what you could do with the dataset at hand. You do not need to adhere to these, it is completely up to you what you would like to try out!

---

### Read in the Dataset:

When you have chosen a dataset for analysis the first thing is to read it into R. If you use one of our datasets or data in an excel sheet you can do this the way we did before, using the `read_excel()` function from `readxl`. 

If you are using your own dataset you will have to figure out which read function you need to get your data into R. There is a function for reading in *almost* any type of text file, `read.csv()`, `read.table()`, `read.delim()`, etc. You can either google your way to an answer or ask one of the course instructors. 

Name your dataset something logical.  

---

### Exploratory Analysis

Before making any statistical models it is necessary to have a good understanding of the dataset.   
You could start by looking at:

- How many variables and observations are there in the data? Which data types are the variables?

- Are there any missing values (`NA`), and how many? If you don't know how to do this we recommend that you google it. You will find many useful answers!  

- Make some preliminary plots of your data. These do not have to be beautiful, but they should be informative to you. You could try histograms, boxplots, barplots or something else you would like.   

- Looks at your plots, do you appear to have any outlier data points in your dataset?   

- Get some summary statistics for your variables, i.e. means, medians, sd, sums (for numerical types) and counts per group (for categorical or factor types).

- Plot the summary stats as well if you think it makes sense.

---

### Scientific Question

Now that you have an overview of your dataset and the variables in it, try to think about and formulate a simple scientific hypothesis/question which you would like to explore. 

Lets use an example of a dataset containing information on patients with breast cancer. Lets say that we follow these patients from diagnosis, though treatment, and 10 years after treatment ends, as follow-up:   

&nbsp;

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)

BC_dataset <-  tibble(
  Age = c("60", "51", "58", "..."),
  BC.subtype = c("luminal", "basal", "luminal", "..."),
  Treatment = c("chemotherapy", "chemotherapy", "immunotherapy", "..."),
  Survival = c("alive", "alive", "dead", "..."),
  Stage = c("2", "3", "4", "..."),
  Smoker = c("no","no", "yes", "..."),
  Familial.BC = c("no", "yes", "no", "..."))

knitr::kable(BC_dataset)
```


&nbsp;
We need to decide on:


- What is our **response/outcome variable (y)**. Using the example dataset above this might be patient survival or patient subgroup (BC.subtype).   
**Q:** Which variable in your chosen dataset will you use as the response variable? And what type is it, e.g. a categorical (two or more discrete types of outcome) or a continuous variable?   

- Decide on our **predictive/explanatory variable(s) (x)**. These are measurements or information such as i.e. age, smoking status or the treatment we want to use to explain or predict the outcome. Some variables like age might modulate the response but not be directly explanatory. We often call these covariates. We do not know whether they truly are covariates or actually explanatory, but using our previous knowledge we hypothesize that these variables influence the effect of treatment on survival. 
**Q:** Which predictive variable(s) (x) do you have in your datatset? Do you have variable(s) you would like to add as covariates?

---

### Data Wrangling

After familiarizing yourself with your dataset and deciding on which variables may be of interest to include in your analysis you might consider whether you need to do any data wrangling/management before defining your statistical model. Here we mean things like:

- Do you need all variables in the dataset? Maybe some are redundant or uninteresting so you could remove them from the tibble.   

- If you have missing values for some observations, do you want to filter out these rows before continuing analysis? There could also be other filtering criteria to think about, such as removing rows with zeros or rows containing measurements that you think are outliers.   

- Maybe you want to make new columns combining some existing ones? An example of this, could be as we did with the downloads dataset where you converted sizes from bytes to megabytes and added the result as a new column to your tibble.   

- Do you want to rename the observations of a variables to something more suited for modeling? An examples of this could be converting a categorical variable like status of smoking (noted as _smoker_ or _non-smoker_) to integer values (_1_ or _0_).

---

### Modelling

Now you have a *tidy* dataset ready for analysis. You have decided on your response and predictive variables and you know which datatypes they are. The latter is important as this will determine which type of model you can make.  

In the statistics exercise we tried out the `lm()` function to do one-way ANOVA. Use this function again to make a simple statistical model.   
If the outcome variable you are interested in is continuous, then you will be doing linear regression. On the other hand if the outcome is categorical, you are performing ANOVA (if you only have two groups, this will effectively be a t-test). **N.B** the function and annotation in R is the same for these models.
Have a look a the statics exercise/presentation and the pseudo code below:

```{r, eval=FALSE}
model1 <- lm(resp ~ pred1, data=name_of_dataset)
```

- Begin with a simple model, use just one predictive variable.   

- Test whether your data fulfills the model assumptions by making residual plots and evaluating these, just like in the presentation and exercise:
```{r, eval=FALSE}
par(mfrow=c(2,2)) # makes room for 4=2x2 plots.
plot(model1)
```

- If model assumptions are not appropriate could you perform some sort of transformation to help alleviate the problem?   

- Get some summary statistics from your model. Can you extract the confidence intervals of your estimate? HINT: `summary()`, `confint()`.     

- Try making a model with more than one predictive variable. Remake the residual plots for you new more complex model.   

- Which model is best? How could you test this? HINT: `drop1()`, `anova()`.   
