---
title: "I. Working with data in R (exercises)"
author: "Science Data Lab & Center for Health Data Science"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  #pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Basic R commands

6.  Load the `trees` dataset by copying and executing the following commands:
    
```{r}
data(trees)
trees
?trees
```

7. Scatter plot of tree Volume as a function of tree height:

```{r}
par(mfrow=c(2,2))
plot(Volume~Height, data=trees)
plot(Volume~Height, data=trees, log="xy")
plot(log(Volume)~log(Height), data=trees)
```    
    
10. Install the tidyverse package:    
```{r, eval=FALSE}
install.packages(tidyverse)
```

11. Load the tidyverse package:    
```{r}
library(tidyverse)
```


12. Median of variable x:
```{r}
x <- c(1,3,8,9,100,NA)
x
median(x)
```

13. Function help and median without NAs
```{r}
?median
median(x, na.rm=TRUE)
```
\

14. Extract variables `Girth, Height, and Volume` from the `trees` dataset and get some summary statistics:   

Extract variables:
```{r}
height <- trees$Height
volume <- trees$Volume
girth <- trees$Girth
```

Summary statistics:
```{r}
length(height)
sum(height)
median(height)
max(height)
min(height)

mean(volume)
var(volume)
sd(volume)
quantile(volume)

summary(trees)

```
\

15. Bonus questions. How do you extract only elements/values `2 to 5` from the `Girth` variable in the trees dataset?

```{r}
girth[2:5]
girth[c(2,3,4,5)]
```
\

    
