---
title: "I\. Working with data in R (solution)"
author: "datalab"
date: '2022-05-20'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic R commands

8.  To start with, load the `tidyverse` package.
    
```{r}
library(tidyverse)
```

9.  Now, load the `trees` dataset by copying and executing the following commands:
    
```{r}
data("trees")
view(trees)
?trees
```

10.  Check what information is contained in the `trees` dataset by calling the `summary` function on it:
    
```{r}
summary(trees)
```

11.  How many trees do we have data for?

```{r}
dim(trees)
```

We have 31 rows, so 31 observations.
    
12.  Now, extract the column called `volume` from the dataset and assign it to a new variable called `volume_col`.

```{r}
volume_col <- trees$Volume
```

13.  Display the volume of the first 10 trees.

```{r}
volume_col[1:10]
```
    
14.  Find the minimum, maximum and mean volume from the column you just extracted. Does it match what was stated in the `summary`?

```{r}
min(volume_col)
max(volume_col)
mean(volume_col)
```

Yes, the values match what was written in `summary` for the volumn column.
    
15.  What class of datastructure is `trees`? Make it into a tibble.
```{r}
class(trees)
tib_trees <- as_tibble(trees)
```

    
