---
title: "Graphics with ggplot2 (solution)"
author: "Data Science Lab, University of Copenhagen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = TRUE, warning = FALSE)
```
 
Loading the core tidyverse packages and the 'readxl' package for data import from .xlsx.

```{r}
library(tidyverse)
library(readxl)
```

Importing the climate data from **climate.xlsx**^[Contains public sector information licensed under the [Open Government Licence v3.0](http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/).]. (Change the path to the Excel file below so that it matches the path to the file saved on your own computer, or use *Import Dataset* in RStudio to obtain the relevant code.)

```{r}
climate <- read_excel("climate.xlsx")
climate
```

### Scatter plot I


1. Make a scatter plot of **rain** against **sun**. HINT: Try `geom_point`.


```{r}
ggplot(climate, aes(x = sun, y = rain)) + 
  geom_point()
```

2. Colour the points in the scatter plot according to weather station. Save the plot in an object, for example `p`.


```{r}
p <- ggplot(climate, aes(x = sun, y = rain, colour = station)) + 
  geom_point()
p
```

3. Add the command `facet_wrap(~station)` to the saved plot object from above, and update the plot. New instructions are added with `+`, see the ggplot lecture. What happens?


```{r}
p + facet_wrap(~station)
```


4. Is it necessary to have a legend in the faceted plot? You can remove the legend by adding `theme(legend.position = "none")` to your code.

```{r}
p <- ggplot(climate, aes(x = sun, y = rain, colour = station)) + 
  geom_point() +
  facet_wrap(~station) +
  theme(legend.position = "none")
p
```


### Scatter plot II: labels

5. Now, let's label each point with the month it corresponds to. This is done by using the geom `geom_text`. 

```{r}
ggplot(climate, aes(x=rain, y = sun, color = station)) +
  geom_point() + facet_wrap(~station) + theme(legend.position = "none") +
  geom_text(aes(label = month)) 
```

6. You might find that your labels are on top of your points and not very readable. Let's put them a bit higher by adding `nudge_y = 15` to the geom_text.

```{r}
ggplot(climate, aes(x=rain, y = sun, color = station)) +
  geom_point() + facet_wrap(~station) + theme(legend.position = "none") +
  geom_text(aes(label = month), nudge_y = 15)
```

7. Change `geom_text` to `geom_label`.  

```{r}
ggplot(climate, aes(x=rain, y = sun, color = station)) +
  geom_point() + facet_wrap(~station) + theme(legend.position = "none") +
  geom_label(aes(label = month), nudge_y = 15)
```

### Graphic files

```{r}
ggsave(file="weather.jpeg")

ggsave(file="weather.png",width=10,height=8,units="cm")
```


### Scatter plot III: error bars

10. Calculate the average and standard deviation for sunshine in each month and save it to a table called `summary_stats`. You will need group_by and summarize.

```{r}
summary_stats <- climate %>%
  group_by(month) %>% 
  summarize(sun_avg = mean(sun), sun_sd = sd(sun))
```

11. Make a scatter plot of the summary_stats with month on the x-axis, and the average number of sunshine hours on the y-axis.

```{r}
p <- ggplot(summary_stats, aes(x = month, y = sun_avg)) +
  geom_point()
p
```


12. Add error bars to the plot, which represent the average number of sunshine hours plus/minus the standard deviation of the observations. The relevant geom is called `geom_errorbar`.

```{r}
p <- ggplot(summary_stats, aes(x = month, y = sun_avg)) +
  geom_point() +
  geom_errorbar(aes(ymin = sun_avg - sun_sd, ymax = sun_avg + sun_sd), width = 0.3)
p
```

13. Could you have made a plot with horizontal error bars instead? How?

```{r}
p + coord_flip()
```

### Line plot

14. Make a line plot of the rainfall observations over time, such that observations from the same station are connected in one line. If you have trouble with this, have a look at the ggplot lecture. Put month on the x-axis. Colour the lines according to weather station as well.

```{r}
ggplot(climate, aes(x = month, y = rain, group = station, colour = station)) +
  geom_line()
```

15. Change month to a factor. Make the line plot again.
```{r}
climate <- mutate(climate, month = factor(month))

ggplot(climate, aes(x = month, y = rain, group = station, colour = station)) +
  geom_line()

```

 
16. Use `theme(legend.position = "top")` to move the colour legend to the top of the plot.

```{r}
ggplot(climate, aes(x = month, y = rain, group = station, colour = station)) +
  geom_line() + theme(legend.position = "top")
```

### Layering

We can add several geoms to the same plot to show several things at once.

17. Use `geom_point()` to add the monthly rainfall data points to the line plot above.

```{r}
ggplot(climate, aes(x = month, y = rain, group = station, colour = station)) +
  geom_line() +
  geom_point() +
  theme(legend.position = "top") 
```

18. Now, insert `geom_hline(yintercept = mean(climate$rain), linetype = "dashed")` at the end of your code for the line plot, and update the plot again. What have you just added to the plot?

```{r}
ggplot(climate, aes(x = month, y = rain, group = station, colour = station)) +
  geom_line() +
  geom_point() +
  theme(legend.position = "top") +
  geom_hline(yintercept = mean(climate$rain), linetype = "dashed")
```

19. Labels:
```{r}
ggplot(climate, aes(x = month, y = rain, group = station, colour = station)) +
  geom_line() +
  geom_point() +
  theme(legend.position = "top") +
  geom_hline(yintercept = mean(climate$rain), linetype = "dashed") +
  labs(x = "Month", y = "Rainfall (mm)", colour = "Weather station")
```


### Box plot I

22. Make a box plot of the sunshine observations by weather station. 

```{r}
ggplot(climate, aes(x = station, y = sun)) +
  geom_boxplot()

```


23. Colour the boxes according to weather station. 


```{r}
ggplot(climate, aes(x = station, y = sun, fill = station)) +
  geom_boxplot()
```

### Box plot II - Aesthetics

24. Move the legend with `theme(legend.position="top")` and add a different legend title with `labs(fill = "Custom Title")`.  

```{r}
ggplot(climate, aes(x = station, y = sun, fill = station)) +
  geom_boxplot() +
  theme(legend.position="top") +
  labs(fill = "Weather Station")
```

25. Change the theme of the ggplot grid. Suggestions: `theme_minimal()`, `theme_bw()`, `theme_dark()`, `theme_void()`.   

```{r}
ggplot(climate, aes(x = station, y = sun, fill = station)) +
  geom_boxplot() +
  theme(legend.position="top") +
  labs(fill = "Weather Station") +
  theme_bw()
```

26. Instead of the colours automatically chosen when you use `fill = station`, pick your own colors. Use the `scale_fill_manual()`. 
You will need five colors, one for each station. What happens if you choose too few colours? 

```{r}
ggplot(climate, aes(x = station, y = sun, fill = station)) +
  geom_boxplot() +
  theme(legend.position="top") +
  labs(fill = "Weather Station") +
  theme_bw() +
  scale_fill_manual(values = c("darkcyan","steelblue", "lightblue", "lavender", "pink"))
```

27. Change the boxplot to a **violin plot**.
    Add the sunshine observations as scatter points to the plot. 
    Include a boxplot inside the violin plot, use `geom_boxplot(width=.1)` to do this.


```{r}
ggplot(climate, aes(x=station, y=sun, fill=station)) + 
  #Change the boxplot to a violin plot
  geom_violin(trim=FALSE) + 
  #Add the sunshine observations as scatter points to the plot. 
  geom_point(position = position_jitter(seed = 1, width = 0.1)) + 
  #Include a boxplot inside the violin plot
  geom_boxplot(width=.1) + 
  scale_fill_manual(values = c("darkcyan","steelblue", "lightblue", "lavender", "pink")) + 
  labs(fill = "Weather station") + theme(legend.position="top") + theme_bw()
```

### Histogram

```{r}
ggplot(climate, aes(x = rain)) +
  geom_histogram(binwidth = 25, colour = "red", fill = "white")
```

### Bar chart I

32. Make a bar chart which visualizes the sunshine hours per month.

```{r}
ggplot(climate, aes(x=as.factor(month), y=sun)) +
  geom_col()

```

33. Colour, i.e. divide the bars according to weather station.

```{r}
ggplot(climate, aes(x = as.factor(month), y = sun, fill = station)) +
  geom_col()
```

34. For better comparison, place the bars for each station next to each other instead of stacking them.

```{r}
ggplot(climate, aes(x=as.factor(month), y=sun, fill = station)) +
  geom_col(position = 'dodge')
```

35. Make the axis labels and legend title of the plot more informative by customizing them like you did for the line plot above.

```{r}
ggplot(climate, aes(x=as.factor(month), y=sun, fill = station)) +
  geom_col(position = 'dodge') + 
  labs(x = "Month", y = "Sunshine (hours)", colour = "Weather station")
```


### Bar chart II: Sorting bars

39. Make a new bar chart showing the annual rainfall recorded at each weather station.

```{r}
ggplot(climate, aes(x = station, y = rain)) +
  geom_col()
```


40. Sort the stations in accordance to rainfall, either ascending or descending. This was shown in the ggplot lecture. 

```{r}
annual_rain <- 
  climate %>%
  group_by(station) %>% 
  summarize(rain = sum(rain)) %>% 
  arrange(rain)

climate <- mutate(climate, station = factor(station, levels = annual_rain$station))

ggplot(climate, aes(x = station, y = rain)) +
  geom_col()

```


41. Add labels to each bar that state the sum of the rainfall. You can do this by using the summarized dataframe created above and passing it to `geom_text` as data, together with a suitable aes mapping. I.e. if you named it `annual_rain` you can write:

```{r}

ggplot(climate, aes(x = station, y = rain)) +
  geom_col() +
  geom_label(mapping = aes(x = station, y = rain, label = rain), data = annual_rain)

```

42. Adjust the label positions so that the labels are positioned immediately above the bars.


```{r}
ggplot(climate, aes(x = station, y = rain)) +
  geom_col() +
  geom_label(mapping = aes(x = station, y = rain, label = rain), data = annual_rain, nudge_y = 60)

```
