---
title: "ggplot lecture - solutions"
author: "datalab"
date: '2022-05-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

0. load libraries and import data

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
downloads <- 
  read_excel("../Presentations/downloads.xlsx") %>% 
  filter(size > 0)
downloads
```


## Exercise A: 10 mins

1. Make a bar chart of the downloads data showing the total download size per month. 
Hint: Very similar to the first example shown during the lecture 

```{r}
ggplot(downloads,aes(x=month,y=size)) +
  geom_col()
```

2. Make a bar chart of the downloads data showing the total time spend on downloads per month. 

```{r}
ggplot(downloads,aes(x=month,y=time)) +
  geom_col()
```

3. Assign the plot you made in 1. to the variable p_size_month .

```{r}
p_size_month <- ggplot(downloads,aes(x=month,y=size)) +
  geom_col()
```

## Exercise B: 7 mins

1. On the bar chart you made in A3 (p_size_month), add coloring by the machineName by using the 'fill' keyword in the aes.

```{r}
p_size_month <- ggplot(downloads,aes(x=month,y=size, fill = machineName)) +
  geom_col()
p_size_month
```

2. Now, position the bars for the different machines next to each other instead of stacked. 
Hint: Use the 'position' keyword.

```{r}
p_size_month <- ggplot(downloads,aes(x=month,y=size, fill = machineName)) +
  geom_col(position = 'dodge')
p_size_month
```

3. Now turn it into a boxplot instead. If it's hard to see the boxes try to make the scale of the size axis logarithmic.

```{r}
p_size_month <- ggplot(downloads,aes(x=month,y=size, fill = machineName)) +
  geom_boxplot() + scale_y_log10()
p_size_month
```

## Exercise C

0. Create daily_downloads dataframe (from lecture).

```{r}
daily_downloads <- downloads %>%
    group_by(machineName, date) %>% 
    summarize(dl_count = n(), size_mb = sum(size)/10^6) %>%
    mutate(total_dl_count = cumsum(dl_count))

daily_downloads
```


1. Add coloring by the total download count (total_dl_count) to this plot:
p <- ggplot(daily_downloads, aes(x = date, y = dl_count)) +
  geom_point()
  
```{r}
p <- ggplot(daily_downloads, aes(x = date, y = dl_count, color = total_dl_count)) +
  geom_point()
p
```

2. Add a different point shape depending on the machine to the same plot.

```{r}
p <- ggplot(daily_downloads, aes(x = date, y = dl_count, color = total_dl_count, shape = machineName)) +
  geom_point()
p
```

3. Change the coloring to be discrete instead of continuous. You can choose total_dl_count > 5000 or any cutoff you like. 

```{r}
p <- ggplot(daily_downloads, aes(x = date, y = dl_count, color = total_dl_count > 5000, shape = machineName)) +
  geom_point()
p
```

