---
title: "Tidyverse lecture - Solutions"
author: "datalab"
date: '2022-05-20'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise A

1. Make a new R script
2. Load the climate data which you can find in the Exercises folder

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
climate <- read_excel("climate.xlsx")
climate
```

3. Is your data a dataframe or a tibble?
```{r}
class(climate)
```

It's a tibble.

4. Have a look at:

4.1 How many observations are there?
```{r}
dim(climate)
```

4.2 How many data columns are there and what are their types?
```{r}
summary(climate)
```


4.3 How many different stations are there?
```{r}
distinct(climate,station)
```


4.4 How many observations per station?
```{r}
climate %>% count(station)
```

## Exercise B: 5-10 mins

On the climate dataset, select:

1. all rows from the station in Oxford

```{r}
climate %>%
  filter(station == 'oxford')
```

2. all rows from the station in Oxford when there were at least 100 hours of sunlight
```{r}
climate %>%
  filter(station == 'oxford' & sun >= 100)
```


3. all rows from the stations in Oxford and Camborne when there were at least 100 hours of sunlight 
```{r}
climate %>%
  filter(station %in% c('oxford','camborne') & sun >= 100)
```

4. a subset that only contains the station, year and rain columns
```{r}
sub <- climate %>%
  select(station, year, rain)
sub
```


## Exercise C

1. To the climate dataset, add:

1.1 A column that states the amount of hours with no sunshine for each month. A month has on average 730 hours, you can use the same amount of hours of all of them.

```{r}
climate <- climate %>%
  mutate(no_sun = 730 - sun)
```

1.2. A column the says whether the weather this month was good. We consider good a month with at least 100 hours of sunlight and less than 100 mm of rain. Otherwise the weather was bad.

```{r}
climate <- climate %>%
  mutate(good_weather = ifelse(sun > 100 & rain < 100, "Yes", "No"))
climate
```


2. Count the number of: 

2.1 Months, i.e. lines, per station that did not have any days with air frost (so two conditions)

```{r}
climate %>%
  count(station, af == 0)
```


2.2 Months with good weather per station (use the column you made in 1.2). What's the place with the best weather in England? 

```{r}
climate %>%
  filter(good_weather == 'Yes') %>%
  count(station) %>%
  arrange(n)
```

Oxford has the most months with good weather (9).


