---
title: "II\. Working with data in R (solution)"
author: "Data Science Lab, University of Copenhagen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = TRUE, warning = FALSE)
```

Loading the core **tidyverse** packages, as well as the **readxl** package for importing data from .xlsx

```{r}
library(tidyverse)
library(readxl)
```

### Importing data

1-3. Use the _Import data facility_. 

```{r}
library(readxl)
climate <- read_excel("climate.xlsx")
climate
```

### Working with the data

```{r}
library(tidyverse)
```

3. Assigning the data set of observations from Oxford with 0 days of air frost to `oxford_af`.

```{r}
oxford_af <- climate %>%
  filter(station == "oxford", af == 0)
oxford_af
```

4. Compute the average rainfall over all stations and months by using the summarize function

```{r}
climate %>%
  summarize(avg_rain = mean(rain))
```

5. Calculate average rainfall, the standard deviation of the monthly rainfall as well as the total rainfall nad put them in the same table. 

```{r}
rain_stats <- climate %>%
  summarize(avg_rain = mean(rain),
          sd_rain = sd(rain),
          sum_rain = sum(rain))
```

6. Summary statistics for rainfall data grouped by weather station.

```{r}
climate %>%
  group_by(station) %>% 
  summarize(avg_rain = mean(rain),
            sd_rain = sd(rain),
            sum_rain = sum(rain))
```

7. Including number of observations for each station.

```{r}
climate %>%
  group_by(station) %>% 
  summarize(avg_rain = mean(rain),
            sd_rain = sd(rain),
            sum_rain = sum(rain),
            n=n())
```
    
8. Sorted in descending order according to annual rainfall.    
 
```{r}
climate %>%
  group_by(station) %>% 
  summarize(avg_rain = mean(rain),
            sd_rain = sd(rain),
            sum_rain = sum(rain),
            n=n()) %>% 
  arrange(desc(sum_rain))
```
 
9. Sorted by average monthly sunshine

```{r}
climate %>%
  group_by(station) %>% 
  summarize(avg_rain = mean(rain),
            sd_rain = sd(rain),
            sum_rain = sum(rain),
            avg_sun = mean(sun),
            n=n()) %>% 
  arrange(avg_sun)
```


    
10. Weather station with largest median number of monthly sunshine hours over the months April to September was Oxford:

```{r}
climate %>% 
  filter(month %in% 4:9) %>% 
  group_by(station) %>% 
  summarize(med_sun = median(sun)) %>% 
  arrange(desc(med_sun))
```

11. For each weather station apart from Armagh, the total rainfall (in cm) and duration of sunshine (in days) in the months with no days of air frost.

```{r}
climate %>% 
  group_by(station) %>% 
  filter(station != "armagh", af == 0) %>% 
  summarize(total_rain = sum(rain)/10,
            total_sun = sum(sun)/24)
```

12. For each month: Number of stations with at least two days of air frost or more than 95 mm rain, and the average sunshine duration for these stations.
    
```{r}
climate %>%
  filter(af >= 2 | rain > 95) %>% 
  group_by(month) %>% 
  summarize(count = n(), avg_sun = mean(sun))
```
    