---
title: "VI. Bioinformatics in R (presentation)"
author: "Center for Health Data Science, University of Copenhagen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  #pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Bioconductor

[Bioconductor](https://www.bioconductor.org/) provides tools for computational biology and bioinformatics analysis in R - it is open source and open development and it has an active user community.   
Mostly when we install R-packages we use `install.packages('name_of_package')`. When we use this command we refer to the [CRAN repository](https://cran.r-project.org/) of packages, however sometimes we want a package from `Bioconductor` instead. For this we use the command `BiocManager::install('name_of_package')`. In order to use this installer, you need to download the R-package `BiocManager` e.g. `install.packages('BiocManager')`.  



## Gene Expression Analysis in R with `DEseq2`

[DEseq2](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) is one of the many packages/frameworks which exists for analysis of bulk gene expression data in R. For more information on DEseq2, please have a look at the original publication [here](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8).   

Other highly used packages for differential expression analysis _DEA_ are:   

* [limma](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)   

* [edgeR](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)   

* [NOIseq](https://www.bioconductor.org/packages/release/bioc/vignettes/NOISeq/inst/doc/NOISeq.pdf)   

`DEseq2` has many advantages over classical models and post hoc tests, as it is specifically developed for handling common issues and biases in expression data, including differences in sequencing depth and highly variable dispersion of counts between genes.   
In brief, `DEseq2` fits a generalized linear model (GLM) for each gene in the dataset. In the case where we compare two groups i.e. treatment vs control, the GLM fit returns coefficients indicating the overall expression strength of a gene, along with the log-2 fold change between groups. `DEseq2` adjusts variable gene dispersion estimates using an empirical Bayes approach which borrows information across genes and shrinks gene-wise dispersions towards a common dispersion trend to increase accuracy of differential expression testing.

&nbsp;

---


### About the Dataset

The dataset used for this presentation was acquired from the following github tutorial on RNAseq analysis: <https://combine-australia.github.io/RNAseq-R/06-rnaseq-day1.html>.

RNA sequencing data generated from luminal and basal cell sub-populations in the mammary gland of three groups of mice:   
- Control   
- Pregnant   
- Lactating

The objective of the original study (found [here](https://pubmed.ncbi.nlm.nih.gov/25730472/)) was to identify genes specifically expressed in lactating mammary glands, the gene expression profiles of luminal and basal cells from different developmental stages were compared.

&nbsp;

---

### Load R-packages:

```{r}
# Data Wrangling
# install.packages("tidyverse")
# install.packages("readxl")
library(tidyverse)
library(readxl)

# For Plotting
# install.packages("ggplot2")
library(ggplot2)


# For DEA
# install.packages("BiocManager")
# BiocManager::install("DESeq2")
library(BiocManager)
library(DESeq2)
```


&nbsp;

---

### Importing Data

Reading in data:
```{r}
exprDat <- read_excel("MouseRNAseq.xlsx")
exprInfo <- read_excel("MouseSampleInfo.xlsx")

# Look at the data:
head(exprDat, n=5)
dim(exprDat)

head(exprInfo)
```
Convert character columns to factor types:
```{r}
exprInfo <- exprInfo %>%
  mutate(CellType = as.factor(CellType),
         Status = factor(Status, levels = c("control", "pregnant", "lactate")),
         Status.Type = as.factor(Status.Type))

head(exprInfo)
```


&nbsp;

---

### Initial Data Check & Filtering:

Let's try to sample 12 (n) random genes and plot their count distribution.

```{r}
expr12 <- exprDat %>%
  dplyr::select(-EntrezGeneID, -GeneName) %>%
  sample_n(.,12) %>% 
  t() %>%
  as_tibble() %>% 
  rename_at(vars(names(.)), ~paste0("Gene", seq(1:12))) %>% 
  gather() %>% 
  mutate(valuelog2 = log2(value+1))

# Give it a look:
expr12
```

Plot:
```{r}
ggplot(expr12, aes(valuelog2)) + 
  geom_histogram(color="black", fill="grey80", bins=20) + 
  theme_minimal() +
  facet_wrap(~key)
```

---

We will filter out genes with too many zero counts. Exclude the columns with gene information:

```{r}
# Count number of 0s across samples. Filter samples where at least four samples has a count great than 0.

exprDat <- exprDat %>% 
  mutate(nzeros = rowSums(dplyr::select(.,-EntrezGeneID, -GeneName)==0)) %>%
  filter(nzeros <= 8) %>%
  dplyr::select(-nzeros)

#How many genes do we have left:
dim(exprDat)
```

&nbsp;

---

## Differential Expression Analysis- DESeq2


We will now make a DESeq2 object. For this we use the function `DESeqDataSetFromMatrix` from the DEseq2 package. As input we give our count matrix, our gene IDs and our meta data (exprInfo). Additionally we include a design for DE contrasts. In this case we add CellType (luminal or basal) and Status (control, pregnant or lactating).   


Convert to exprDat to a dataframe and make GeneNames column into rownames:
```{r}
# Pull out GeneNames and EntrezGeneID for later use
GeneNames <- exprDat %>%
  dplyr::select(EntrezGeneID, GeneName)

  
exprDat <- exprDat %>%
  dplyr::select(-EntrezGeneID) %>%
  column_to_rownames(., var = "GeneName")
```


Make a DESeq2 object:
```{r}
exprObj <- DESeqDataSetFromMatrix(countData = exprDat,
                              colData = exprInfo,
                              design= ~CellType+Status)
exprObj
```

---

Next, we use `DEseq()` to estimate dispersion, gene-wise and mean-dispersion, fitting model(s):
```{r}
exprObj <- DESeq(exprObj)
```


---


### Preliminary analysis:

Let's have a look at the library sizes:

```{r}
colSums(assay(exprObj))
```

The count distributions may be dominated by a few genes with very large counts. These genes will drive plotting e.g. heatmaps, PCA analysis etc.
Let's see if we have any genes with high large counts and in turn, dispersion in our dataset. For convenience I am using the base R boxplot function:

```{r}
#boxplot(assay(exprObj), las=2)
boxplot(log2(assay(exprObj)+1), las=2)
```

We perform variance stabilizing transformation to obtain log2 counts per million read mapped, overcoming issues with outlier genes and sequencing depth:

```{r}
exprObjvst <- vst(exprObj,blind=FALSE)
boxplot(assay(exprObjvst), xlab="", ylab="Log2 counts per million",las=2)
```

&nbsp;

---

### Principal Component Analysis

Before performing DEA it is a good idea to explore how samples cluster together based on there gene expression profile. The expectation here is that samples from the same group (treatment vs control, condition A vs condition B, etc.) will cluster together. A principal component analysis (PCA) plot can also help us to identify outlier samples which might need to be removed from the analysis. We use our vst counts for principal component analysis:

```{r}
plotPCA(exprObjvst,intgroup=c("Status"))
plotPCA(exprObjvst,intgroup=c("CellType"))
#plotPCA(exprObj,intgroup=c("TypeStatus"))
```

&nbsp;

---

### Testing

Have a look at the group comparisons: 

```{r}
resultsNames(exprObj)
```

&nbsp;

---

Test for DE genes between the three groups of mice, adjusted for cell type:   

(I) lactating and control mice:
```{r}
resLC <- results(exprObj, contrast = c("Status", "lactate", "control"), independentFiltering = FALSE)
```

Summary and plot of DE analysis results:

```{r}
DESeq2::plotMA(resLC)
summary(resLC)
```


Custom function to filter results of DEA. We make a function to save writing the same code three times in a row, one time for each comparison:   

```{r}
# SIGNIFICANT DE GENES:
# Takes as arguments:
# my.res = a dataframe of results from the DEseq results()function
# my.LFC = log fold change cutoff, default is 1.0
# my.cof = adjusted p-value cutoff, default is 0.01 
SigDE <- function(my.res) {
  my.res <- as.data.frame(my.res) %>%
    rownames_to_column(., var = "GeneName") %>%
    as_tibble() %>%
    mutate(dir = ifelse(log2FoldChange >= 0, 'up', 'down')) %>%
    filter((log2FoldChange >= 1.0 | log2FoldChange <= -1.0) & padj <= 0.01) %>%
    arrange(padj, desc(abs(log2FoldChange)))
  return(my.res)
}

```
   
   
Filter DEA results from comparison of lactating vs control mice using custom function:   

```{r}
resLC <- SigDE(resLC)

# Number of DE genes:
dim(resLC)

# Give it a look
head(resLC, n=5)
```

---

Below we perform the same steps as above to get the DE genes between (II) pregnant and control mice and (III) lactating and pregnant mice:  

(II) pregnant and control mice:
```{r}
resPC <- results(exprObj, contrast = c("Status", "pregnant", "control"),  independentFiltering = FALSE)
#DESeq2::plotMA(resPC)
#summary(resPC)
resPC <- SigDE(resPC)

# Number of DE genes:
dim(resPC)
```

(III) lactating and pregnant mice:
```{r}
resLP <- results(exprObj, contrast = c("Status", "lactate", "pregnant"),  independentFiltering = FALSE)
#DESeq2::plotMA(resLP)
#summary(resLP)
resLP <- SigDE(resLP)

# Number of DE genes:
dim(resLP)
```



&nbsp;

---

### Heatmap Visualization

To visually inspect if DE genes identified in our DESeq2 analysis successfully separate the three groups of mice (control, pregnant and lactating), we will make a heatmap. 
For this we use the `heatmap` function and package `viridis`.

It will not make sense to include all DE genes in this heatmap (3000 genes). Instead pick the top 50 most significant DE genes, based on adj. p-value and logFC.

---

Make a vector of unique EntrezGeneIDs (top50):
```{r}
topDE <- bind_rows(resPC[1:50,], resLC[1:50,], resLP[1:50,]) %>%
  pull(GeneName) %>%
  unique()

length(topDE)
```

---

The expression counts themselves (not logFC) are needed for the heatmap. We use the topDE vector to extract these from the vst normalized DESeq2 object.

```{r}
head(assay(exprObjvst), n=5)

resVST <- assay(exprObjvst) %>% 
  as.data.frame() %>%
  rownames_to_column(var = "GeneName")  %>%
  as_tibble() %>%
  filter(GeneName %in% topDE)
```


The heatmap function in base R wants gene expression data as a matrix (a dataframe with numeric values only). We extract the GeneNames column and convert the tibble into a matrix:
```{r}
HPnames <- resVST %>% 
  pull(GeneName)


HPdat <- resVST %>%
  dplyr::select(-GeneName) %>%
  as.matrix()
```


---

We use the `heatmap` function to generate a heatmap. We can modify the look of the heatmap as desired, e.g. add column colors, row labels, change color scheme etc.

```{r}
heatmap(HPdat, labRow=FALSE,
        ColSideColors=exprInfo$CellType.colors, 
        labCol=exprInfo$Status, 
        cexCol=1.2, 
        cexRow = 1.3)
```



